SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2012-12-10 paper/paper.tex
done with reduction algorithm description.

\title{SmartCheck}
%\subtitle{Subtitle Text, if any}

@@

\begin{abstract}
foobar
\end{abstract}

@@

\begin{figure}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline 
 & QC (none) & QC (10) & QC (20) & SmartCheck \\      
\hline \hline 
Max.  & 0.15s & 21.51s & 125.37s & 1.96s\\
\hline
Mean  & 0.07s & 1.21s & 3.80s & 0.30s\\
\hline
Median & 0.07s & 0.48s & 0.52s & 0.24s\\
\hline
Mean size & 70 & 34 & 32 & 5\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 100 tests of the program in Figure~\ref{fig:initial}
    using interpreted Haskell.}
  \label{fig:results}
\end{figure}

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Figure~\ref{fig:results}.  The table characterizes testing the program shown
in Figure~\ref{fig:initial} that we have described so far.  Four trials are
presented.  In the first three, we use QuickCheck to generate counterexamples.
The first column contains the results of using QuickCheck without shrinking to
provide a baseline.  The next two columns show the result of using QuickCheck
with definitions of \ttp{shrink} that limit the lists of fields of \ttp{B} to 10
(Figure~\ref{lst:newshrink}) and 20 elements (Figure~\ref{lst:newshrink}, with
\ttp{len = 20}), respectively.  (Recall that with the original definition for
shrink from Figure~\ref{fig:initial}, some counterexamples cause QuickCheck to
not return after hours of computation.)  In each row, we show the maximum, mean,
and median number of seconds required by QuickCheck to generate a counterexample
over 100 runs.  In the last row, we show the mean size of the final value
returned, measured by the taking the cumulative length of the lists of the
fields of \ttp{B}.  Notice that while typically counterexamples are returned in
a fraction of a second, there are outliers taking 21 seconds and 125 seconds for
the two tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

@@

\subsection{Reduction Algorithm Overview}
The algorithm we present for efficiently searching for new counterexamples can
be thought of as a recursive breadth-first traversal over the structure of the
current counterexample value, where the value is represented as a tree.  At each
node, during the traversal, we generate arbitrary \emph{structurally} smaller
sub-values and build a new value from that, leaving the remainder of the tree
unchanged.  Intuitively, by a structurally smaller value, we mean one with fewer
constructors.  We continue until we reach a fixed-point.

@@

\paragraph{Bounding counterexample exploration}
\todo{limit depth}
\todo{limit size of generated structures}

@@

%% For example, consider the following types that give us a binary tree
%% %
%% \begin{code}
%% data T = N Int
%%        | B T T
%% \end{code}
%% %
%% \noindent
%% and suppose that the property is that the sum of the leaves is less than 10.
%% Starting from the following counterexample
%% %
%% \lstset{caption={A counterexample value.}
%%        ,label=lst:val}
%% \begin{lstlisting}
%% B (B (N 2) (N 3)) 
%%   (B (N 4) (B (N 5) (N 6)))
%% \end{lstlisting}
%% %
%% \noindent
%% Consider the sub-value \ttp{B (N 4) (B (N 5) (N 6))}.  Some example values of
%% structurally smaller well-typed values include the following:
%% %
%% \begin{code}
%% N 2
%% B (N 8) (N 7)
%% B (N 0) (N 1)
%% \end{code}
%% %
%% \noindent
%% Replacing the sub-value, we get the following new values, each of which is
%% smaller than the value in Listing~\ref{lst:val}:
%% %
%% \begin{code}
%% B (B (N 2) (N 3)) 
%%   (N 2)

%% B (B (N 2) (N 3)) 
%%   (B (N 8) (N 7))

%% B (B (N 2) (N 3)) 
%%   (B (N 0) (N 1))
%% \end{code}

In this work, we limit ourselves to algebraic data types.

\todo{GADTs?}

\section{Counterexample Generalization}

@@

@@

%--------------------------------------------------------------------------------

%--------------------------------------------------------------------------------
\section{Related Work}
\label{sec:related}

@@

%--------------------------------------------------------------------------------
\section{Conclusions}
\label{sec:conclusions}

>>>>

\title{SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization}
%\subtitle{Subtitle Text, if any}

@@

\begin{abstract}
QuickCheck is a powerful library for automatic test-case generation.  Because
QuickCheck performs random testing, it can discover very large counterexamples,
even when enumerating test-cases is infeasible, as is done with SmallCheck.
QuickCheck provides an interface for the user to write a \emph{shrink} function
to attempt to reduce the size of a counterexample discovered by QuickCheck.
Implementations of \emph{shrink} often require significant boilerplate and are
not ``smart'', just generating structurally smaller values.  We introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples.  In addition to shrinking, SmartCheck generalizes
counterexamples by returning a formula that abstracts over individual
counterexamples.  SmartCheck has been implemented for Haskell and tested on a
variety of Haskell programs.
\end{abstract}

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline 
 & QC (none) & QC (10) & QC (20) & SmartCheck \\      
\hline \hline 
Max.  & 0.15s & 21.51s & 125.37s & 1.96s\\
\hline
Mean  & 0.07s & 1.21s & 3.80s & 0.30s\\
\hline
Median & 0.07s & 0.48s & 0.52s & 0.24s\\
\hline
Mean size & 70 & 34 & 32 & 5\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 100 tests of the program in Figure~\ref{fig:initial}
    using interpreted Haskell.}
  \label{table:results}
\end{table}

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Table~\ref{fig:results}.  The table characterizes testing the program shown
in Figure~\ref{fig:initial} that we have described so far.  Four trials are
presented.  In the first three, we use QuickCheck to generate counterexamples.
The first column contains the results of using QuickCheck without shrinking to
provide a baseline.  The next two columns show the result of using QuickCheck
with definitions of \ttp{shrink} that limit the lists of fields of \ttp{B} to 10
(Figure~\ref{lst:newshrink}) and 20 elements (Figure~\ref{lst:newshrink}, with
\ttp{len = 20}), respectively.  (Recall that with the original definition for
shrink from Figure~\ref{fig:initial}, some counterexamples cause QuickCheck to
not return after hours of computation.)  In each row, we show the maximum, mean,
and median number of seconds required by QuickCheck to generate a counterexample
over 100 runs.  In the last row, we show the mean size of the final value
returned, measured by the taking the cumulative length of the lists of the
fields of \ttp{B}.  Notice that while typically counterexamples are returned in
a fraction of a second, there are outliers taking 21 seconds and 125 seconds for
the two tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

@@

\subsection{Reduction Algorithm Overview}\label{sec:reduct}
The algorithm we present for efficiently searching for new counterexamples can
be thought of as a recursive breadth-first traversal over the structure of the
current counterexample value, where the value is represented as a tree.  At each
node, during the traversal, we generate arbitrary \emph{structurally} smaller
sub-values and build a new value from that, leaving the remainder of the tree
unchanged.  Intuitively, by a structurally smaller value, we mean one with fewer
constructors.  We continue until we reach a fixed-point.

@@

@@

The number of constructors in an algebraic data value primarily determines how
large and understandable it is.  In our experience, the size of fields that come
from primitive types, like \ttp{Int} or \ttp{Char}, have less of a role in
understanding a counterexample.

Consequently, we distinguish a class of types referred to as \emph{base types}.
If the reduction algorithm encounters a base type, it is ignored.  By default,
this includes numeric types (e.g., \ttp{Int}, \ttp{Word}, \ttp{Float}, etc.,
\ttp{Bool}, and \ttp{Char}.  We also treat lists specially: lists of base types
are considered to be base types themselves.  This prevents the algorithm from
analyzing, for example, strings.  A list of a non-base-types is shrunk by the
algorithm.

By providing custom instances, the user can declare any substructure in a data
type to be a base type.  Doing so can help make SmartCheck more efficient if the
user knows that some portion of the structure cannot be shrunk or is irrelevant
to the property.

\paragraph{Bounding counterexample exploration}

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxFailure :: Int
scMaxSize    :: Int
scMaxDepth   :: Maybe Int  
\end{code}
%
\noindent
The ``\ttp{sc}'' prefix stands for \emph{S}mart\emph{C}heck and is used to
prevent name clashes with QuickCheck's flags.  \ttp{scMaxFailure} controls the
number of values generated by the reduction algorithm for each sub-value
analyzed.  \ttp{scMaxSize} controls the maximum size of values generated to
replace sub-values by the reduction algorithm.  Thus, new sub-values must be
strictly smaller than the minimum of \ttp{scMaxSize} and the size of the
sub-value being replaced.  (The size of the value is with respect to the
\ttp{size} method in the \ttp{SubTypes} class.)  Finally, \ttp{scMaxDepth}
determines the maximum depth in the counterexample the reduction algorithm
should analyze.  A value of \ttp{Nothing} means that the counterexample should
be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) = 1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

\section{Counterexample Generalization} \label{sec:generalization}
Suppose we are testing a value.

@@

\todo{GADTs? don't work}

@@

\todo{xmonad}

%-------------------------------------------------------------------------------

%-------------------------------------------------------------------------------
\section{Related Work}
\label{sec:related}

@@

%-------------------------------------------------------------------------------
\section{Conclusions}
\label{sec:conclusions}
